%!TEX root = ../template.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% chapter2.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\typeout{NT FILE chapter2.tex}%

\chapter{Background}
\label{cha:background}

%%------------------------------------------------------------
%% SECTION: Foundations of Neural Language Representation
%%------------------------------------------------------------
\section{Foundations of Neural Language Representation}
\label{sec:foundations}

The ability to represent language in a form that machines can process and reason
over is a prerequisite for any modern Generative AI system. Two landmark
contributions established the theoretical and practical foundations upon which
contemporary large language models (LLMs) and retrieval-augmented architectures
are built: the introduction of dense word representations via neural networks,
and the development of the Transformer architecture based solely on attention
mechanisms.

\subsection{Distributed Word Representations}
\label{sec:word2vec}

Prior to 2013, natural language processing relied predominantly on sparse,
high-dimensional representations such as one-hot encodings, which offered no
intrinsic notion of semantic similarity between words. Mikolov et
al.~\cite{mikolov2013word2vec} addressed this limitation by proposing
Word2Vec, a computationally efficient framework for learning continuous vector
representations of words from large unlabelled corpora. The model is trained
through a self-supervised prediction task using one of two architectures:
Continuous Bag-of-Words (CBOW), which predicts a target word from its
surrounding context, or Skip-gram, which performs the inverse prediction.
The resulting word embeddings encode both syntactic and semantic structure
in a dense vector space, enabling linear algebraic operations over meaning ---
most notably the well-known relationship \(\vec{king} - \vec{man} + \vec{woman}
\approx \vec{queen}\).

The significance of Word2Vec extends well beyond the paper itself. By
demonstrating that semantic similarity can be captured as geometric proximity
in a continuous vector space, it established the distributional hypothesis as
an actionable engineering principle. This insight directly underlies the design
of modern embedding models and vector databases, which are central components
in Retrieval-Augmented Generation (RAG) pipelines~\cite{lewis2020rag}. In the
context of this work, the quality and cost of the embedding stage in the
Maestro framework's RAG pipeline is a direct consequence of choices made in
this representational paradigm.

\subsection{The Transformer Architecture}
\label{sec:transformer}

Despite the advances in distributed representations, sequence modelling
remained constrained by recurrent architectures such as LSTMs and GRUs, which
process tokens sequentially and suffer from vanishing gradients over long
contexts. Vaswani et al.~\cite{vaswani2017attention} proposed a radical
departure: the Transformer, an architecture that dispenses with recurrence
entirely and relies solely on attention mechanisms to model dependencies between
all positions in a sequence simultaneously. The core operation, scaled
dot-product attention, computes output representations as a weighted sum of
value vectors, where weights are derived from the compatibility between query
and key projections:

\[
\text{Attention}(Q, K, V) = \text{softmax}\!\left(\frac{QK^{T}}{\sqrt{d_k}}
\right)V
\]

By running multiple such attention operations in parallel --- Multi-Head
Attention --- the model learns to simultaneously attend to different types of
relationships across the input, such as syntactic dependencies and semantic
co-references. Positional encodings are injected additively into the input
embeddings to compensate for the architecture's inherent permutation
invariance. The resulting model achieved 28.4 BLEU on the WMT 2014
English-to-German translation benchmark, surpassing all prior approaches while
requiring a fraction of the training compute~\cite{vaswani2017attention}.

The Transformer's parallelism over sequence positions unlocked a new scaling
regime for neural language models. Architectures derived from it --- most
notably encoder-only models such as BERT~\cite{devlin2018bert} for
representation learning, and decoder-only models such as the GPT
series~\cite{brown2020gpt3} for text generation --- became the standard
building blocks of enterprise GenAI systems. Critically, the attention
mechanism scales quadratically with sequence length, \(O(n^2)\), meaning
that context length is a primary cost driver in token-priced LLM APIs.
This architectural property is foundational to the FinOps strategies
explored in this dissertation, namely prompt compression, semantic caching,
and dynamic model selection, as each seeks to reduce the effective \(n\)
processed by the model or the frequency with which inference is invoked at
all~\cite{mikolov2013word2vec,vaswani2017attention}.

Together, these two works define the representational substrate of modern
GenAI: Word2Vec established that meaning can be encoded as position in a
learned vector space, and the Transformer established that context-sensitive
reasoning over such representations can be performed efficiently and at scale.
The remainder of this chapter examines the architectural patterns, operational
frameworks, and cost optimisation techniques that build upon these foundations
in production-grade enterprise systems.

%%------------------------------------------------------------
%% SECTION: Large Language Models in Production
%% Covers: LLM overview and scaling laws, RLHF/instruction tuning,
%% open-weight models, and the enterprise deployment spectrum.
%%------------------------------------------------------------
\section{Large Language Models in Production}
\label{sec:llms-in-production}

The preceding section established the two foundational breakthroughs that made
modern language AI possible: distributed word representations and the
Transformer architecture. Building upon these primitives, the field witnessed
a rapid consolidation around a unified paradigm — the large language model
(LLM) — in which ever-larger Transformer-based networks, pre-trained on
internet-scale corpora, acquire general-purpose language competence that can be
redirected toward downstream tasks with minimal additional supervision. This
section characterizes LLMs as a class, traces the methodological evolution from
raw pre-training to instruction-following alignment, and examines the
architectural and economic implications of deploying such systems at enterprise
scale.

\subsection{Large Language Models: An Overview}
\label{sec:llm-overview}

% Defines LLMs via the emergent-capability criterion; introduces decoder-only
% architecture and contrasts with encoder-only BERT; establishes scaling laws
% as the link between compute investment and token-cost motivation.
A large language model is a neural language model of sufficient parameter count
and training data scale that general linguistic and reasoning capabilities emerge
as a consequence of the pre-training objective alone, without task-specific
architectural modification~\cite{brown2020gpt3}. Pre-training proceeds via
self-supervised next-token prediction over corpora of hundreds of billions to
trillions of tokens, optimizing the model to assign high probability to
naturally occurring text. The resulting parameter weights encode an implicit
representation of syntax, semantics, world knowledge, and rudimentary reasoning
chains that can be elicited through natural language prompting.

LLMs are predominantly realized as \textit{decoder-only} Transformer
architectures~\cite{vaswani2017attention}, in which the self-attention layers
apply a causal mask that restricts each token's attention to its leftward
context, preserving the autoregressive generation property. This design choice
contrasts with the encoder-only variant typified by BERT~\cite{devlin2018bert},
which produces bidirectional contextual representations optimized for
discriminative tasks. For generative applications — intelligent assistants,
automated reasoning, and document synthesis — the decoder-only paradigm
dominates, and it is this class of model that underpins the enterprise GenAI
systems examined in this dissertation.

A defining empirical finding in the development of LLMs is the existence of
\textit{scaling laws}: systematic, power-law relationships between model size,
training compute, dataset volume, and downstream performance~\cite{brown2020gpt3}.
Scaling laws imply that performance is largely predictable and improvable by
increasing any of the three axes, establishing a framework within which
investment in compute translates into measurable capability gains. This property
has driven the successive scaling of commercial models and, critically, has also
made token consumption a dominant cost variable: larger models process the same
input more expensively, directly motivating the cost optimization strategies
central to this work.

\subsection{From Pre-training to Instruction Tuning}
\label{sec:instruction-tuning}

% Motivates the alignment gap between next-token pre-training and user intent;
% describes the three-stage RLHF pipeline (SFT → reward model → PPO) from
% InstructGPT; highlights the key empirical result that a 1.3B aligned model
% outperforms a 175B unaligned one — directly informing dynamic model selection.
Despite the generative competence acquired during pre-training, early LLMs
exhibited a systematic misalignment between model behavior and user intent.
A model trained solely to predict the next token on a web-scale corpus learns to
continue text in statistically plausible ways, but not necessarily in ways that
are \textit{helpful}, \textit{honest}, or \textit{harmless} when deployed as an
interactive assistant. Ouyang et al.~\cite{ouyang2022instructgpt} addressed this
gap by introducing a three-stage fine-tuning pipeline centered on
Reinforcement Learning from Human Feedback (RLHF).

The procedure operates as follows. First, a \textit{supervised fine-tuning}
(SFT) phase trains the base model on high-quality demonstrations of desired
behavior, produced by human labelers on a curated prompt distribution. Second, a
\textit{reward model} is trained to predict human preference scores over pairs of
model outputs, internalizing a proxy for human judgment. Third, the SFT model is
further optimized against the reward model signal using Proximal Policy
Optimization (PPO), an online reinforcement learning algorithm. The resulting
InstructGPT models, despite comprising as few as 1.3 billion parameters, were
consistently preferred by human evaluators over the unaligned 175-billion
parameter GPT-3 baseline, demonstrating that alignment quality is not strictly
a function of scale~\cite{ouyang2022instructgpt}. This result has profound
practical implications: it established RLHF as the standard post-training
procedure for production LLMs and confirmed that an aligned smaller model can
match or exceed the utility of a larger unaligned one — a relationship that
informs the dynamic model selection strategies explored in this
dissertation.

\subsection{Open-Weight Models and the Deployment Spectrum}
\label{sec:open-weight-models}

% Contrasts proprietary API models (GPT-4, Claude) with self-hostable
% open-weight alternatives (Llama 2); introduces the LLM cascading / dynamic
% model selection paradigm as the core cost-optimization mechanism evaluated
% in the Maestro framework.
The commercial LLM landscape is stratified between proprietary, API-accessed
models and openly released, self-hostable alternatives. The former class,
exemplified by OpenAI's GPT-4 family and Anthropic's Claude series, exposes
inference through token-priced cloud APIs; the latter is represented most
prominently by Meta AI's Llama series. Touvron et al.~\cite{touvron2023llama2}
introduced Llama~2, a family of models ranging from 7 to 70 billion parameters,
pre-trained on two trillion tokens and released under a license permitting
commercial use. Fine-tuned variants — Llama~2-Chat — further apply RLHF
alignment, achieving performance competitive with closed-source alternatives on
standard benchmarks while remaining locally deployable~\cite{touvron2023llama2}.

The existence of capable open-weight models introduces a consequential degree of
freedom into enterprise GenAI architectures. Rather than committing uniformly to
a single proprietary provider, organizations may route inference dynamically
across a portfolio of models differentiated by capability tier, latency
characteristics, and per-token cost. Smaller open-weight models, deployed
on-premise or on reserved cloud instances, can serve high-volume, lower-complexity
queries at substantially reduced marginal cost, while API-accessed frontier
models handle tasks requiring peak reasoning performance. This model routing
paradigm — often referred to as \textit{LLM cascading} or
\textit{dynamic model selection} — is a central cost optimization mechanism
evaluated in this dissertation in the context of the Maestro framework's
multi-cloud deployment~\cite{touvron2023llama2, ouyang2022instructgpt}.

Together, the progression from raw pre-training through RLHF alignment to the
open-weight deployment spectrum defines the operational environment within which
enterprise GenAI frameworks must function. The ability to select, combine, and
switch between models of varying capability and cost is not merely an
architectural convenience but a financial imperative in production systems
where token consumption can scale three to five times per client without
deliberate optimization~\cite{ouyang2022instructgpt}.

%%------------------------------------------------------------
%% SECTION: Related Work
%%------------------------------------------------------------
\section{Related Work}
\label{sec:related-work}

% TODO: Compare and contrast existing work with your approach.
% Organise by sub-topic if appropriate.


%%------------------------------------------------------------
%% SECTION: Summary
%%------------------------------------------------------------
\section{Summary}
\label{sec:background-summary}

% TODO: Summarise the main takeaways from the literature and position your work.
